Variational theory of non-relativistic quantum electrodynamics

### 2018-10-16 paper4pg.tex

\title{Variational theory of the ground state of quantum electrodynamics}

@@

\begin{abstract}
The past decade has brought rapid theoretical and experimental advances in the field of ultra-strong coupling of light and matter from microwave to optical frequencies. Many of the theoretical analyses of such systems proceed by considering simple two-level systems coupled to a single cavity mode. There remains a need to understand light-matter coupling  in contexts of complicated electronic systems coupling to many-mode or continuum photonic systems, as well as a need to understand how vacuum fluctuations of electromagnetic fields are altered in real space by strong light-matter coupling in such complex systems.  In this work, we develop a variational theory of the ground state of a quantum electrodynamical system of coupled light and matter. Essential to our ansatz is the notion an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling. Our variational formulation leads to a set of general equations that can describe the ground state of multi-electron systems coupled to many photonic modes in real space. As a first step towards an  \textit{ab initio} approach to ground states in QED, we apply our ansatz to a Rabi model involving a multi-level emitter with many optical modes, a model which cannot be analytically solved. We find a compact semi-analytic formula which describes the ground state energy very accurately (to less than 1\% error) in all regimes of coupling parameters allowed by the Thomas-Reiche-Kuhn sum rule. We expect our method to give rise to highly accurate descriptions of ground-state phenomena in quantum electrodynamics in complicated systems, such as Casimir forces, Lamb shifts, and fluctuational electrodynamics phenomena. 
\end{abstract}

@@

The past decade has brought an explosion of effort and experimental progress in realizing the coupling of matter and quantized electromagnetic fields beyond a regime of weak coupling, in which perturbative treatments of quantum electrodynamical couplings can no longer accurately describe the dynamics \cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong,kockum2018ultrastrong}. Ultra-strong, or even deep-strong coupling is now regularly observed in superconducting qubits at microwave frequencies \cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, ensembles of molecules in optical cavities \cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, 2D electron gas cyclotron resonances coupled to cavities \cite{hagenmuller2010ultrastrong,scalari2012ultrastrong,zhang2016collective}, intersubband transitions in quantum wells coupled to cavities \cite{todorov2010ultrastrong,geiser2012ultrastrong}, and single-emitters in ultra-confined plasmonic geometries \cite{benz2016,chikkaraddy2016}. Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, highly ionized atoms coupled to optical media via the Cerenkov effect, and many more. The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems \cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties \cite{orgiu2015}, realizing analogues of nonlinear optical processes with vacuum fluctuations \cite{kockum2017deterministic}, and even as new sources of photons through the dynamical Casimir effect \cite{ciuti2005quantum}.

In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved. In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model~\cite{braak2011}, which describes the coupling of a two-level atom to a single mode, the Hopfield model \cite{hopfield1958theory}, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators \cite{dicke1954coherence}. Any realistic system which deviates from these simpler models require numerical diagonalization approaches for an exact solution, which become unwieldy when the dimension of the electronic Hilbert space becomes large, as is the case when describing an electronic system in real space. It also becomes unwieldy when one needs to keep track of many cavity modes which can also be populated with many photons. Exciting new developments towards ab-initio studies of QED have proceeded by proposing reduced-quantity descriptions such as density-functional descriptions of the light-matter coupling \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}. With such density-functional descriptions, workers in this field have been able to compute ground-state and time-dependent properties of realistic molecules in optical cavities \cite{flick2017c}.

In what follows, we propose the use of variational methods for the solution of \emph{ab initio} QED problems.  In particular, we develop an ansatz in which the ground state can be considered as a factorizable state of an effective matter and effective photon quasiparticle both in their respective ground states. This ansatz, reminiscent to the Hartree-Fock ansatz of electronic structure theory, leads to coupled eigen-equations describing the ground-state of the light-matter system. We benchmark our ansatz against a multi-level and multi-mode Rabi model parameterized by the strength of the characteristic momentum matrix elements describing the coupling of the matter and the electromagnetic fields. We find that for light-matter couplings that respect the Thomas-Reiche-Kuhn sum rule, our theory finds the ground state energy to a remarkable accuracy of more than >99\%, even in non-perturbative coupling regimes.  In regimes where our results are accurate, we claim we have found the effective quasiparticle description of the ground state. Moreover, in this regime, the ground state interaction energy manifests itself as the difference in zero-point energies between the quasiparticles and the bare particles.  Our findings furnish a non-perturbative theory of the Lamb shift, allowing us to non-perturbatively calculate position-dependent Lamb shifts of ground and excited states that give rise to Casimir-Polder forces.

@@

The QED Hamiltonian takes the form $H = H_{mat}+H_{em}+H_{int}$ where $H_{mat}$ describes the matter in the absence of the quantized electromagnetic field, $H_{em}$ describes the photons in the absence of the matter, and $H_{int}$ describes the coupling between light and matter. The matter Hamiltonian takes the form:
\begin{align}
H_{el} &= \int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext}(x) \right)\psi(x) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
\end{align}
where $v_{ext}$ is the one-body potential, $V(x-x')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field. %\Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}

Parameterizing the electromagnetic field purely in terms of a vector potential: $\mathbf{E} = -\partial_t\mathbf{A}$ and $\mathbf{B} = \nabla\times\mathbf{A}$. This renders the free electromagnetic Hamiltonian as
\begin{equation}
H_{em} = \int d^3x~ \epsilon (\partial_t \mathbf{A}(x))^2 + \mathbf{A}(x)\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(x)),
\end{equation}
where $\epsilon$ and $\mu$ represent a non-dispersive and positive dielectric and magnetic background that the matter and photon occupy. For cases we consider in this work, these will be taken to be unity, but we leave them here for the sake of generality.

@@

This Hamiltonian, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions $\{\psi_n\}$, and in terms of a set of photonic mode functions $\{\mathbf{F}_i\}$ via mode expansions for the electronic and photonic field operators. The electron field operator takes the form $\psi(x) = \sum_n \psi_n(x)c_n$.
In this second-quantized description, the $c_n$ is an annihilation operator for an electron corresponding to state $n$. The electromagnetic field operator takes the form $\mathbf{A}(x) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(x)a_i+\mathbf{F}^*_i(x)a^{\dagger}_i\right)$. In the electromagnetic field operator, we parameterize not only by the mode functions but also by modal frequencies. We have chosen a form for the vector potential operator which mirrors the choice generally adopted in QED. The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}.$

With the Hamiltonian set up, we now move to describe a variational theory of the ground state. In the variational theorem, we choose some ansatz $|\Omega\rangle$ for the ground state of $H$. The variational theorem ensures that $\langle \Omega|H|\Omega\rangle$ is an upper bound for the ground state energy. Parameterizing the ground state to generate a family of ground states, and minimizing $\langle \Omega|H|\Omega\rangle$ with respect to the introduced parameters gives the best upper bound for the ground state energy for the chosen family of ground states.  We choose as our ansatz
\begin{equation}
|\Omega\rangle = \left( \prod\limits_n c_n^{\dagger}|0_n\rangle,\right) \otimes \left(\bigotimes_i|0_i\rangle\right),
\end{equation}
In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons. Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions and photon mode functions. Thus, we shall minimize the expectational value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$. Note that the electromagnetic field gets two sets of (distinct) variational parameters, $\omega_i$ and $\mathbf{F}_i$. The last thing to note before taking the expectation values and minimization is that we should focus on the submanifold of variations that keep the complete set of electron and photon functions normalized. This shall be enforced by constructing the Lagrange function:
\begin{widetext}
\begin{equation}
\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle - \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right),
\end{equation}
\end{widetext}
with the $E_n$ and $\lambda_i$ being the Lagrange multipliers that enforce the normalization condition. Evaluating the expectation value of the Hamiltonian, and minimizing the Lagrange function  with respect to the orbitals, photon modes, and photon frequencies, immediately yields:
\begin{align}
&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi^*_j(x')\psi_j(x')\psi_i(x) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi_j(x')\psi_j(x)\psi_i(x')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)   = E_i\psi_i(x).
\end{align}
for the electron orbitals and energies. One sees that in addition to the one-body and Hartree-Fock terms for the electrons, the effect of the QED coupling is to add a one-body ponderomotive potential. We note that in the often used dipole approximation, the ponderomotive potential has no effect as it becomes independent of position.
\begin{figure}[t]
\includegraphics[width=8.5cm]{conceptfigure.pdf}
\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons. (Right) Quasiparticle description of the coupled system as a factorable state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
\label{fig:ansatz}
\end{figure}
For the photon orbitals and energies, the minimization yields:
\begin{equation}
\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2}{\omega_q^2} \right)\right)\mathbf{F}_q = 0,
\end{equation}
where $\omega_p^2 = \frac{e^2}{m\epsilon_0}\sum\limits_{i=1}^N |\psi_i(x)|^2$ is a position-dependent squared-plasma frequency which will push the photon orbitals out of the region where the emitter is located.

Immediately, one notices that the term in the interaction Hamiltonian linear in the vector potential makes no contribution to the expectation value of the ground state of the energy in this ansatz. In many quantum electrodynamical problems, this linear term, which has zero expectation value in our ansatz, is important. To first order, it leads to spontaneous emission of a single photon. To second order, it leads to two (virtual) photon processes, such as van der Waals / Casimir-Polder forces on emitters, real photon processes such as two-photon emission, and effective interactions betweeen distinct emitters. Thus, we seek to capture the effect of this term. Physically, this term will mix the factorizable ground state of Equation (7) with terms that have virtual excitations of the matter, as well as virtual excitations of the electromagnetic field. The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of the quantum electrodynamical ground state.
\begin{figure}[t]
\includegraphics[width=5cm]{figure2new.pdf}
\caption{\textbf{Variational theory of ground and excited states in the ultra-strong coupling regime of QED.} Lowest two eigenvalues as a function of coupling for the multi-mode Rabi model describing the coupling of an emitter to a one-dimensional cavity. The variational results (blue) agree to within roughly a percent with the result derived by numerical diagonalization (orange) retaining up to four photons in the Hilbert space. The result from perturbation theory (green) disagrees strongly with the other methods of calculation for the first excited state due to a near-resonance between the bare transition energy and the bare lowest cavity mode.}
\label{fig:ansatz}
\end{figure}
We capture the effect of correlations perturbatively. In other words, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential. That correction is given by 
\begin{equation}
\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{q=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_q^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_q(\omega_{mn} -\lambda_q)}
\end{equation}
where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.  In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (6) and (7), and then the electron energies and orbitals as well as the photon frequencies and orbitals, are plugged into Equation (8).
\begin{figure}[t]
\includegraphics[width=8.4cm]{figure3new.pdf}
\caption{\textbf{Variational theory of ground state energy: accuracy as a function of number of emitter levels.} (Left) Ground-state energy of the one-dimensional multi-mode Rabi model as a function of the dimensionless coupling for two (top), three (middle), and four (bottom) level emitter systems with momentum matrix elements chosen to saturate the Thomas-Reiche-Kuhn sum rule. (Right) Ratio of the energy calculated with the variational theory and with numerical diagonalization. These results show that our accuracy is robust to the number of levels in the system. }
\label{fig:ansatz}
\end{figure}
In what follows, we provide a proof-of-concept demonstration of the efficacy of the variational theory derived here. We consider the QED Hamiltonian corresponding to a single emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction. As the cavity is considered for simplicity to be one-dimensional, the electric field is oriented along a single direction, denoted $x$, while the magnetic field is oriented along a direction transverse to both the electric field and the cavity length, denoted $y$.

>>>>

\title{Variational theory of non-relativistic quantum electrodynamics}

@@

\begin{abstract}
The almost routine ability to achieve ultra-strong coupling between light and matter promises to bring about new means to control  material properties, new concepts for manipulating light at the atomic scale, and fundamentally new insights into quantum electrodynamics (QED). Thus, there is a need to develop quantitative theories of QED phenomena in complex electronic and photonic systems. In this letter, we develop a variational theory of general non-relativistic QED systems of coupled light and matter. Essential to our ansatz is the notion an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling. This variational formulation leads to a set of general equations that can describe the ground state of multi-electron systems coupled to many photonic modes in real space. As a first step towards an  \textit{ab initio} approach to ground states in QED, we apply our ansatz to a QED system involving a multi-level emitter coupled to many optical modes. We find a compact semi-analytic formula which describes ground and excited state energies to less than 1\% error in all regimes of coupling parameters allowed by sum rules. Moreover, our formulation provides essentially a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as "the Casimir energy of a single atom in a cavity". Our method should give rise to highly accurate descriptions of ground-state phenomena in general QED systems, such as Casimir forces, Lamb shifts, and other fluctational electrodynamical phenomena.
\end{abstract}

@@

Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED) \cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong,kockum2018ultrastrong}. Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits \cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules \cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, Landau level systems \cite{hagenmuller2010ultrastrong,scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities \cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems \cite{benz2016,chikkaraddy2016}. Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect, and many more. The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems \cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties \cite{orgiu2015}, realizing analogues of nonlinear optical processes with vacuum fluctuations \cite{kockum2017deterministic}, and as probes of exotic phenomena such as the dynamical Casimir effect \cite{ciuti2005quantum}. Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}. Through "reduced quantity theories" such as quantum electrodynamical density functional theory (QEDFT) \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}, one is now able to calculate observables in large molecules coupled to realistic optical cavities  \cite{flick2017c}.

In this letter, we propose a variational framework to analyze complex light-matter systems from first principles. Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of functionals to describe the ground state, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the use of the long-wavelength (dipole) approximation. A variational framework, as we shall show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation. Beyond these advantages, a variational framework also allows conceptual insights, as we shall show, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom. A variational framework also allows compact semi-analytic formulas to describe complex systems which may allow the development of functionals for use in QEDFT.
%In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved. In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model~\cite{braak2011}, which describes the coupling of a two-level atom to a single mode, the Hopfield model \cite{hopfield1958theory}, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators \cite{dicke1954coherence}. Any realistic system which deviates from these simpler models require numerical diagonalization approaches for an exact solution, which become unwieldy when the dimension of the electronic Hilbert space becomes large, as is the case when describing an electronic system in real space. It also becomes unwieldy when one needs to keep track of many cavity modes which can also be populated with many photons. Exciting new developments towards ab-initio studies of QED have proceeded by proposing reduced-quantity descriptions such as density-functional descriptions of the light-matter coupling \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}. With such density-functional descriptions, workers in this field have been able to compute ground-state and time-dependent properties of realistic molecules in optical cavities \cite{flick2017c}.

Motivated by all of these potential advantages, we now develop an ansatz in which the ground state can be considered as a factorizable state of an effective matter and effective photon quasiparticle both in their respective ground states. This ansatz $-$ reminiscent to, but qualitatively distinct from, the Hartree-Fock ansatz of electronic structure theory $-$ leads to coupled eigen-equations describing the ground-state of the light-matter system. We apply our ansatz to describe ground and excited states in a multi-level emitter coupled to many photonic modes. We find that for light-matter couplings that respect sum rules, our theory finds the ground state energy to a remarkable accuracy of about 99\%, even in deeply non-perturbative coupling regimes.  In regimes where our results are accurate, we have found the effective quasiparticle description of the ground state of QED.  Our findings also furnish a non-perturbative theory of the position-dependent Lamb shifts of ground and excited states that give rise to Casimir-Polder forces.

@@

The QED Hamiltonian  $H = H_{mat}+H_{em}+H_{int}$ where $H_{mat}$ describes the matter in the absence of the quantized electromagnetic field, $H_{em}$ describes the photons in the absence of the matter, and $H_{int}$ describes the coupling between light and matter. The matter Hamiltonian takes the form:
\begin{align}
H_{el} &= \int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext}(x) \right)\psi(x) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
\end{align}
where $v_{ext}$ is the one-body potential, $V(x-x')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field. %\Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}
\begin{figure}[t]
\includegraphics[width=8.5cm]{conceptfigure.pdf}
\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons. (Right) Quasiparticle description of the coupled system as a factorable state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
\label{fig:ansatz}
\end{figure}
Parameterizing the electromagnetic field purely in terms of a vector potential: $\mathbf{E} = -\partial_t\mathbf{A}$ and $\mathbf{B} = \nabla\times\mathbf{A}$. This renders the free electromagnetic Hamiltonian as
\begin{equation}
H_{em} = \int d^3x~ \epsilon (\partial_t \mathbf{A}(x))^2 + \mathbf{A}(x)\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(x)),
\end{equation}
where $\epsilon$ and $\mu$ represent a non-dispersive and positive dielectric and magnetic background that the matter and photon occupy. For cases we consider in this work, these will be taken to be unity, but we leave them here for generality.

@@

This Hamiltonian, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions (orbitals) $\{\psi_n\}$, and in terms of a set of photonic mode functions (orbitals) $\{\mathbf{F}_i\}$. The electron field operator takes the form $\psi(x) = \sum_n \psi_n(x)c_n$.
The $c_n$ is an annihilation operator for an electron corresponding to state $n$. The electromagnetic field operator takes the form $\mathbf{A}(x) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(x)a_i+\mathbf{F}^*_i(x)a^{\dagger}_i\right)$, where the $a_i^{(\dagger)}$ annihilate (create) a photon in mode $i$. In the electromagnetic field operator, we parameterize not only by the mode functions but also by modal frequencies. The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}.$

With the Hamiltonian described, we now move to develop a variational theory of the ground state. In the variational theorem, we choose some ansatz $|\Omega\rangle$ for the ground state of $H$. The variational theorem ensures that $\langle \Omega|H|\Omega\rangle$ is an upper bound for the ground state energy. Parameterizing the ground state to generate a family of ground states, and minimizing $\langle \Omega|H|\Omega\rangle$ with respect to the introduced parameters gives the best upper bound for the ground state energy for the chosen family of ground states.  We choose as our ansatz
\begin{equation}
|\Omega\rangle = \left( \prod\limits_n c_n^{\dagger}|0_n\rangle,\right) \otimes \left(\bigotimes_i|0_i\rangle\right),
\end{equation}
In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons (see Figure 1). Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions and photon mode functions. Thus, we shall minimize the expectational value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$. Note that the electromagnetic field gets two sets of (distinct) variational parameters, $\omega_i$ and $\mathbf{F}_i$. We enforce that the matter and photon remain normalized by constructing the Lagrange function:
\begin{widetext}
\begin{equation}
\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle - \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right),
\end{equation}
\end{widetext}
with the $E_n$ and $\lambda_i$ being the Lagrange multipliers that enforce the normalization conditions. Evaluating the expectation value of the Hamiltonian, and minimizing the Lagrange function  immediately yields:
\begin{align}
&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi^*_j(x')\psi_j(x')\psi_i(x) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi_j(x')\psi_j(x)\psi_i(x')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)   = E_i\psi_i(x).
\end{align}
for the electron orbitals and energies. We see that in addition to the one-body and Hartree-Fock terms for the electrons, the effect of the QED coupling is to add a one-body ponderomotive potential. We note that in the often used dipole approximation, the ponderomotive potential has no effect as it becomes independent of position.

For the photon orbitals and energies, the minimization yields:
\begin{equation}
\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2}{\omega_i^2} \right)\right)\mathbf{F}_i = 0,
\end{equation}
where $\omega_p^2 = \frac{e^2}{m\epsilon_0}\sum\limits_{n=1}^N |\psi_n(x)|^2$ is a position-dependent squared-plasma frequency which will push the photon orbitals out of the region where the emitter is located.

Immediately, one notices that the term in the interaction Hamiltonian linear in the vector potential makes no contribution to the expectation value of the ground state of the energy in this ansatz. In many quantum electrodynamical problems, this linear term. To second order, it leads to two (virtual) photon processes, such as van der Waals / Casimir-Polder forces on emitters, two-photon spontaneous emission, and effective interactions betweeen distinct emitters. Thus, we seek to capture the effect of this term. Physically, this term will mix the factorizable ground state of Eq. (5) with states that simultaneously have virtual excitations of the matter and the electromagnetic field. The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of the quantum electrodynamical ground state.
\begin{figure}[t]
\includegraphics[width=5cm]{figure2new.pdf}
\caption{\textbf{Variational theory of ground and excited states in the ultra-strong coupling regime of QED.} Lowest two eigenvalues as a function of coupling for the multi-mode Rabi model describing the coupling of an emitter to a one-dimensional cavity. The variational results (blue) agree to within roughly a percent with the result derived by numerical diagonalization (orange) retaining up to four photons in the Hilbert space. The result from perturbation theory (green) disagrees strongly with the other methods of calculation for the first excited state due to a near-resonance between the bare transition energy and the bare lowest cavity mode.}
\label{fig:ansatz}
\end{figure}
We capture the effect of correlations perturbatively. In other words, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential. That correction is given by 
\begin{equation}
\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{i=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{n=1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_i^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_i(\omega_{mn} -\omega_i)}
\end{equation}
where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.  In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Eqs. (6) and (7), and then the electron energies and orbitals as well as the photon frequencies and orbitals, are plugged into Eq. (8).
\begin{figure}[t]
\includegraphics[width=8.4cm]{figure3new.pdf}
\caption{\textbf{Variational theory of ground state energy: accuracy as a function of number of emitter levels.} (Left) Ground-state energy of the one-dimensional multi-mode Rabi model as a function of the dimensionless coupling for two (top), three (middle), and four (bottom) level emitter systems with momentum matrix elements chosen to saturate the Thomas-Reiche-Kuhn sum rule. (Right) Ratio of the energy calculated with the variational theory and with numerical diagonalization. These results show that our accuracy is robust to the number of levels in the system. }
\label{fig:ansatz}
\end{figure}

In what follows, we provide a proof-of-concept demonstration of the efficacy of the variational theory derived here. We consider the QED Hamiltonian corresponding to a single emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction. As the cavity is considered for simplicity to be one-dimensional, the electric field is oriented along a single direction, denoted $x$, while the magnetic field is oriented along a direction transverse to both the electric field and the cavity length, denoted $y$. Working under the long-wavelength or dipole approximation, the Hamiltonian can then be written as:
\begin{equation}\label{eq:hamiltonian}
H = H_{\text{matter}}+\frac{\epsilon_0}{2}\int dz~(E^2+c^2B^2)+\frac{q}{m}A(d)p + \frac{q^2}{2m}A^2(d).
\end{equation}
The fields can be expressed as a mode expansion, where for a cavity of length $L$, the modes are given by $F_n(z) = \sqrt{\frac{2}{L}}\sin\left(\frac{n\pi z}{L} \right)$ and the corresponding mode frequencies are $\omega_n = \frac{n\pi c}{L}$. The matter Hamiltonian we take to be a multilevel system with an arbitrary number of levels, $N_a$. The matter system we describe as an $N_a$ site system. This could be considered as a simple model of a molecule within a tight-binding description. Thus we parameterize the general family of matter Hamiltonians as:
\begin{equation}\label{eq:matter_hamiltonian}
H_{\text{matter}} = \sum\limits_{i=1}^{{N_a-1}} V_i|i\rangle\langle i|+t(|i\rangle\langle i+1|+|i+1\rangle\langle i|) 
\end{equation}
The momentum operator, we write as 
\begin{equation}\label{eq:momentum_operator}
p = \frac{-i\hbar}{R}\sum\limits_{i=1}^{N_a-1} \left(|i\rangle\langle i+1|-|i+1\rangle\langle i| \right),
\end{equation}
where $R$ is a constant with units of length representing roughly the difference in positions between sites. This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the Thomas-Reiche-Kuhn (TRK) sum rule is enforced. In other words: $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1$, where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states.  Since the TRK is based on a full electronic real-space description, this sum rule does not rigorously apply to a discrete-level system. However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot. It thus places a bound on how strong the effect of the $A\Jadd{\cdot}p$ term can be. The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.
